---
title: "data_wrangling"
author: "Nikki Shintaku"
date: "4/16/2020"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
getwd()
library(tidyverse)
library(cowplot)
library(ggthemes)
library(lubridate)

USGS_gage_height <- read.csv("./Data/Raw/USGS_tahoe_gage_height.csv")
NOAA_climate <- read.csv("./Data/Raw/NOAA_Tahoe_climate_data.csv")

#theme
mytheme <- theme_stata(base_size = 14, base_family = "sans", scheme = "s2mono") +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)
```

```{r, include=FALSE}
#changing to date format
USGS_gage_height$datetime <- as.Date(USGS_gage_height$datetime, format = "%m/%d/%y") 
USGS_gage_height$datetime <- format(USGS_gage_height$datetime, "%y%m%d")

#paste 19 if the input is greater than 181231 or 20 if it is less than 
create.early.dates <- (function(d) {
       paste0(ifelse(d > 191231,"19","20"),d)
       })
USGS_gage_height$datetime <- create.early.dates(USGS_gage_height$datetime)
USGS_gage_height$datetime <- as.Date(USGS_gage_height$datetime, format = "%Y%m%d")
class(USGS_gage_height$datetime)
```

```{r, include=FALSE}
#changing date on NOAA Data
class(NOAA_climate$DATE)
NOAA_climate$DATE <- as.Date(NOAA_climate$DATE, format = "%Y-%m-%d")
class(NOAA_climate$DATE)
```

```{r, include=FALSE}
#Wrangling USGS data, keeping only datetime, qual code, and gage_height
#drop NAs and get rid of outliers 1-01-1920 : 4-10-1920
USGS_date_gage <- USGS_gage_height %>%
  select(datetime, gage_height, qual_code) %>%
  drop_na(gage_height) %>%
  filter(datetime > as.Date("1920-04-10"))

names(USGS_date_gage)[1] <- "DATE"

write.csv(USGS_date_gage, file = "./Data/Processed/USGS_Site10337000_gage_processed.csv", row.names = FALSE)
```

```{r, include=FALSE}
#Wrangling NOAA data
#TAHOE CITY, CA US
NOAA_Tahoecity_climate <- 
  NOAA_climate %>%
  filter(NAME == "TAHOE CITY, CA US") %>%
  select(NAME, DATE:TMIN)

interval(first(NOAA_Tahoecity_climate$DATE), last(NOAA_Tahoecity_climate$DATE))
#1903-09-13 UTC--2020-03-30 UTC

write.csv(NOAA_Tahoecity_climate, file = "./Data/Processed/NOAA_TahoeCity_climate_processed.csv", row.names = FALSE)
```

```{r, include=FALSE}
#combining datasets into one
USGS_NOAA_combined <- full_join(USGS_date_gage, NOAA_Tahoecity_climate)

USGS_NOAA_combined <- arrange(USGS_NOAA_combined, DATE)

interval(first(USGS_NOAA_combined$DATE), last(USGS_NOAA_combined$DATE))
class(USGS_NOAA_combined$DATE)

USGS_NOAA_combined <- mutate(USGS_NOAA_combined, TMAX_C = (TMAX - 32) * (9/5))
USGS_NOAA_combined <- mutate(USGS_NOAA_combined, TMIN_C = (TMIN - 32) * (9/5))

write.csv(USGS_NOAA_combined, file = "./Data/Processed/USGS_NOAA_combined_processed.csv", row.names = FALSE)
```

#Data Visualization

```{r, include=FALSE}
library(viridis)
library(RColorBrewer)
library(colormap)

USGS_gage_processed <- read.csv("./Data/Processed/USGS_Site10337000_gage_processed.csv")
NOAA_climate_processed <- read.csv("./Data/Processed/NOAA_TahoeCity_climate_processed.csv")
USGS_NOAA_combined <- read.csv("./Data/Processed/USGS_NOAA_combined_processed.csv")

#theme
mytheme <- theme_stata(base_size = 14, base_family = "sans", scheme = "s2mono") +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
#mytheme <- theme_bw(base_size = 14) +
  #theme(axis.text = element_text(color = "black"), 
       # legend.position = "top",
        #panel.grid.major = element_line(size = 0.2, linetype = 'solid', color = "grey70"),
        #panel.grid.minor = element_blank())

theme_set(mytheme)

USGS_gage_processed$DATE <- as.Date(USGS_gage_processed$DATE, format = "%Y-%m-%d") 
class(USGS_gage_processed$DATE)

NOAA_climate_processed$DATE <- as.Date(NOAA_climate_processed$DATE, format = "%Y-%m-%d")

USGS_NOAA_combined$DATE <- as.Date(USGS_NOAA_combined$DATE, format = "%Y-%m-%d")

```

```{r}
#ggplot(USGS_NOAA_combined, aes(x= DATE, y = gage_height, color = PRCP)) +
  #geom_point(size = 0.5, alpha = 0.5) +
  #scale_color_gradientn(colors = rainbow(5), limits = c(0,0.5), breaks = c(0, 0.25, 0.5))

#tmax_plot <- ggplot(USGS_NOAA_combined, aes(x = DATE, y = TMAX_C)) +
  #geom_point(size = 0.5, alpha = 0.5, shape = 1) +
  #labs(x = "Date", y = "Max Daily Temperature (degree C)")
#print(tmax_plot)

gage_height_plot <- ggplot(USGS_NOAA_combined, aes(x = DATE, y = gage_height)) +
  geom_line(color = "dodgerblue2") +
  labs(x = "Date", y = "Gage Height (ft)") +
  ylim(0, 10) 
#print(gage_height_plot)

PRCP_plot <- ggplot(USGS_NOAA_combined, aes(x = DATE, y = PRCP)) +
  geom_point(shape = 1, size = 0.5, color = "indianred2") +
  labs(x = "Date", y = "Precipitation (in)") +
  ylim(0, 8)
#print(PRCP_plot)

snow_plot <- ggplot(USGS_NOAA_combined, aes(x = DATE, y = SNOW)) +
  geom_point(size = 0.5, shape = 8, color = "plum3", alpha = 0.5) +
  labs(x = "Date", y = "Snow Fall (in)") +
  ylim(0, 45) 
#print(snow_plot)

tmin_plot <- ggplot(USGS_NOAA_combined, aes(x = DATE, y = TMIN_C, fill = TMIN_C > 0)) +
  geom_bar(stat = "identity")+
  labs(x = "Date", y = expression(paste("Minimum Temperature",degree,"C"))) +
  ylim(-70, 50) +
  scale_fill_manual(values = c("blue4", "red2")) +
  theme(legend.position = "none",
        plot.caption = element_text(hjust = 0, face = "italic"))
#print(tmin_plot)

graphs_combined <- plot_grid(gage_height_plot, PRCP_plot, snow_plot, tmin_plot, 
                             labels = "AUTO",
                             align = "h",
                             axis = "b")


#caption <- ggdraw() +
 # draw_figure_label("Figure 1: Lake Tahoe climate data and water level time series. (A) Gage height measurements (lake level) in feet where the lake elevation is measured at 6,220 feet. Lake elevation plus daily gage height measurement will give the changes in lake level. (B) Daily precipitation in Tahoe City, CA. (C) Daily snow fall in Tahoe City, CA. (D) Daily minimum temperatures in Tahoe City, CA.",
            # position = "bottom.left",
           #  x = 0,
            # y =0,
            # size = 10)

#snow_plot <- ggplot(USGS_NOAA_combined, aes(x = DATE, y = SNOW)) +
  #geom_point(size = 0.5, shape = 8, color = "plum3", alpha = 0.5) +
  #labs(x = "Date", y = "Snow Fall (in)", caption = "Figure 1: Lake Tahoe climate data and water level time series. (A) Gage height measurements 
     #  (lake level) in feet where the lake elevation is measured at 6,220 feet. Lake elevation plus 
      # daily gage height measurement will give the changes in lake level. (B) Daily precipitation 
       #in Tahoe City, CA. (C) Daily snow fall in Tahoe City, CA. (D) Daily minimum temperatures in 
       #Tahoe City, CA.") +
 # ylim(0, 45) +
  #theme(plot.caption = element_text(hjust = 0, face = "italic"))

#plot_grid(graphs_combined)    


```

```{r, include=TRUE,fig.width=10, fig.height = 6, fig.cap="Figure 1: Lake Tahoe climate data and water level time series. (A) Gage height measurements (lake level) in feet where the lake elevation is measured at 6,220 feet. Lake elevation plus daily gage height measurement will give the changes in lake level. (B) Daily precipitation in Tahoe City, CA. (C) Daily snow fall in Tahoe City, CA. (D) Daily minimum temperatures in Tahoe City, CA."}

plot_grid(graphs_combined) 


```
Figure 1: Lake Tahoe climate data and water level time series. (A) Gage height measurements (lake level) in feet where the lake elevation is measured at 6,220 feet. Lake elevation plus daily gage height measurement will give the changes in lake level. (B) Daily precipitation in Tahoe City, CA. (C) Daily snow fall in Tahoe City, CA. (D) Daily minimum temperatures in Tahoe City, CA.



