---
title: "GLM"
author: "Nikki Shintaku"
date: "4/21/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
getwd()
library(tidyverse)
library(ggthemes)
library(lubridate)

USGS_gage_processed <- read.csv("./Data/Processed/USGS_Site10337000_gage_processed.csv")
NOAA_climate_processed <- read.csv("./Data/Processed/NOAA_TahoeCity_climate_processed.csv")
USGS_NOAA_combined <- read.csv("./Data/Processed/USGS_NOAA_combined_processed.csv")
USGS_NOAA_samedates <- read.csv("./Data/Processed/USGS_NOAA_samedate_processed.csv")

#theme
mytheme <- theme_stata(base_size = 14, base_family = "sans", scheme = "s2mono") +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)
```

```{r}
#Changing to date format
USGS_gage_processed$DATE <- as.Date(USGS_gage_processed$DATE, format = "%Y-%m-%d") 
class(USGS_gage_processed$DATE)

NOAA_climate_processed$DATE <- as.Date(NOAA_climate_processed$DATE, format = "%Y-%m-%d")

USGS_NOAA_combined$DATE <- as.Date(USGS_NOAA_combined$DATE, format = "%Y-%m-%d")
```

**Single Linear Regression** Continuous response, one continuous explanatory variable.
**Multiple Linear Regression** Continuous response, two or more continuous explanatory variables.

If multiple explanatory variables are chosen, they may be analyzed with respect to their **main effects** on the model (i.e., their separate impacts on the variance explained) or with respsect to their **interaction effects,** the effect of interacting explanatory variables on the model. 

# Assumptions of the GLM

The GLM is based on the assumption that the data residuals approximate a normal distribution (or a linearly transformed normal distribution). We will discuss the non-parametric analogues to several of these tests if the assumptions of normality are violated. For tests that analyze categorical explanatory variables, the assumption is that the variance in the response variable is equal among groups. Note: environmental data often violate the assumptions of normality and equal variance, and we will often proceed with a GLM even if these assumptions are violated. In this situation, justifying the decision to proceed with a linear model must be made. 

## Linear Regression
The linear regression, like the t-test and ANOVA, is a special case of the **generalized linear model** (GLM). A linear regression is comprised of a continuous response variable, plus a combination of 1+ continuous response variables (plus the error term). The deterministic portion of the equation describes the response variable as lying on a straight line, with an intercept and a slope term. The equation is thus a typical algebraic expression: 
$$ y = \alpha + \beta*x + \epsilon $$

The goal for the linear regression is to find a **line of best fit**, which is the line drawn through the bivariate space that minimizes the total distance of points from the line. This is also called a "least squares" regression. The remainder of the variance not explained by the model is called the **residual error.** 

The linear regression will test the null hypotheses that

1. The intercept (alpha) is equal to zero.
2. The slope (beta) is equal to zero

Whether or not we care about the result of each of these tested hypotheses will depend on our research question. Sometimes, the test for the intercept will be of interest, and sometimes it will not.

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

#Single Linear Regression

```{r}
#can we predict lake level (gage height) from precipitation
gage_height_slm <- lm(data = USGS_NOAA_samedates, gage_height ~ PRCP)
summary(gage_height_slm)
#for every 1(in) of precipitation, there is -0.107 gage height. 0(in) in precip is 5.86 ft of gage height
#0.01% of lake level is explained by precip
#significant pvalue 0.01853

cor.test(USGS_NOAA_samedates$gage_height,USGS_NOAA_samedates$PRCP)
#t = -2.355, df = 22732, p-value = 0.01853
# cor -0.0156181
#Gage height decreases with more rain??
```

1. **Residuals vs. Fitted.** The value predicted by the line of best fit is the fitted value, and the residual is the distance of that actual value from the predicted value. By definition, there will be a balance of positive and negative residuals. Watch for drastic asymmetry from side to side or a marked departure from zero for the red line - these are signs of a poor model fit.

2. **Normal Q-Q.** The points should fall close to the 1:1 line. We often see departures from 1:1 at the high and low ends of the dataset, which could be outliers. 

3. **Scale-Location.** Similar to the residuals vs. fitted graph, this will graph the squared standardized residuals by the fitted values. 

4. **Residuals vs. Leverage.** This graph will display potential outliers. The values that fall outside the dashed red lines (Cook's distance) are outliers for the model. Watch for drastic departures of the solid red line from horizontal - this is a sign of a poor model fit.

```{r, fig.height = 3, fig.width = 4}
par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(gage_height_slm)
par(mfrow = c(1,1))
#don't put in a report, just use to validate your results 
```

```{r}
gageheight_by_prcp <- ggplot(USGS_NOAA_samedates, aes(x = PRCP, y = gage_height)) +
  geom_point() +
  geom_smooth(method = "lm")
print(gageheight_by_prcp)
#no linear relationship

ggplot(USGS_NOAA_samedates, aes(x = PRCP, y = log(gage_height))) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
#single lm with log gage height
gage_height_slm2 <- lm(data = USGS_NOAA_samedates, log(gage_height) ~ PRCP)
summary(gage_height_slm2)
#one unit increase in prcp, a change in log value of gage height is -0.026
#p-value: 0.008027, 22732 DF, Adjusted R-squared:  0.0002651

par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(gage_height_slm2)
par(mfrow = c(1,1))
```

```{r}
gage_height_slm3 <- lm(data = USGS_NOAA_samedates, gage_height ~ TMIN)
summary(gage_height_slm3)
#p-value: < 2.2e-16 , Adjusted R-squared:   0.02527 , 22732 degrees of freedom

par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(gage_height_slm3)
par(mfrow = c(1,1))

gage_height_slm3 <- lm(data = USGS_NOAA_samedates, log(gage_height) ~ TMIN)
summary(gage_height_slm3)

ggplot(USGS_NOAA_samedates, aes(x = TMIN, y = gage_height)) +
  geom_point()
```

```{r}
gage_height_slm4 <- lm(data = USGS_NOAA_samedates, gage_height ~ TMAX)
summary(gage_height_slm4)
#p-value: < 2.2e-16 , Adjusted R-squared:  0.01105, 22732 degrees of freedom

par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(gage_height_slm4)
par(mfrow = c(1,1))

gage_height_slm5 <- lm(data = USGS_NOAA_samedates, log(gage_height) ~ TMAX)
summary(gage_height_slm4) #same results as reg data 

ggplot(USGS_NOAA_samedates, aes(x = TMAX, y = gage_height)) +
  geom_point() +
  geom_smooth(method = "lm")
```


```{r}
#non-parametric equivalent: Spearman's Rho
cor.test(USGS_NOAA_samedates$gage_height, USGS_NOAA_samedates$PRCP, method = "spearman", exact = FALSE)
#rho -0.02840093 , S = 2.0139e+12, p-value = 1.845e-05
```

#Multiple Regression

```{r}
gage_height_mlm <- lm(data = USGS_NOAA_samedates, gage_height ~ PRCP + TMIN + SNOW)
summary(gage_height_mlm)
#p-value: < 2.2e-16, TMIN and SNOW and PRCP are significant predictors of gage height 
#Adjusted R-squared:  0.02595  %3 of our data is explained by predictors 
#df 22730 

ggplot(USGS_NOAA_combined, aes(x = PRCP, y = gage_height, color = TMIN)) +
  geom_point()

par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(gage_height_mlm)
par(mfrow = c(1,1))

```

```{r}
#Akaike's information criterion (AIC) step regression
#Tahoe.naomit <- USGS_NOAA_combined %>%
 # select(gage_height, PRCP, SNOW, SNWD, TMAX, TMIN) %>%
  #na.omit()

Gage_AIC <- lm(data = USGS_NOAA_samedates, gage_height ~ PRCP + SNOW + SNWD + TMAX + TMIN)
step(Gage_AIC)
#lm(formula = gage_height ~ PRCP + SNWD + TMAX + TMIN, data = USGS_NOAA_samedates)
#(Intercept)         PRCP         SNWD         TMAX         TMIN  
  #  4.85924     -0.31145      0.01462     -0.02103      0.06739  

Gage_model_AIC <- lm(data = USGS_NOAA_samedates, gage_height ~ PRCP + SNWD + TMAX + TMIN)
summary(Gage_model_AIC)
#p-value: < 2.2e-16, Adjusted R-squared:  0.03838, 22729 DF
#all variables are sig predictors of gage height 
AIC(Gage_model_AIC)
plot(Gage_model_AIC, which = c(1))
plot(Gage_model_AIC, which = c(2))
car::vif(Gage_model_AIC)
lmtest::bptest(Gage_model_AIC)
#ols_plot_cooksd_bar(Gage_model_AIC)
```





