---
title: "data_analysis"
author: "Nikki Shintaku"
date: "4/19/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
getwd()
library(tidyverse)
library(ggthemes)
library(lubridate)
library(trend)
library(zoo)

USGS_gage_processed <- read.csv("./Data/Processed/USGS_Site10337000_gage_processed.csv")
NOAA_climate_processed <- read.csv("./Data/Processed/NOAA_TahoeCity_climate_processed.csv")
USGS_NOAA_combined <- read.csv("./Data/Processed/USGS_NOAA_combined_processed.csv")

#theme
mytheme <- theme_stata(base_size = 14, base_family = "sans", scheme = "s2mono") +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)
```

```{r}
#Changing to date format
USGS_gage_processed$DATE <- as.Date(USGS_gage_processed$DATE, format = "%Y-%m-%d") 
class(USGS_gage_processed$DATE)

NOAA_climate_processed$DATE <- as.Date(NOAA_climate_processed$DATE, format = "%Y-%m-%d")

USGS_NOAA_combined$DATE <- as.Date(USGS_NOAA_combined$DATE, format = "%Y-%m-%d")
```

#Questions for Time Series Analysis

* Has there been an increasing or decreasing **trend** in the response variable over time?
* Can we **forecast** conditions in the future?

Lake water level (gage height) is measured daily at the Tahoe City gage station. Since we are working with one location measured over time, this will make a great example dataset for time series analysis. 

```{r}
ggplot(USGS_gage_processed) + 
  geom_line(aes(x = DATE, y = gage_height))
```

```{r}
#creating time series object on gage height, daily data
Tahoe_gage_height_ts <- ts(USGS_gage_processed[[2]], frequency = 365)
```

```{r}
# Generate the decomposition
Tahoe_gage_height_Decomposed <- stl(Tahoe_gage_height_ts, s.window = "periodic")
plot(Tahoe_gage_height_Decomposed)
```

```{r}
# extracting the components and turn them into data frames to visualize
Tahoe_gage_height_Components <- as.data.frame(Tahoe_gage_height_Decomposed$time.series[,1:3])
Tahoe_gage_height_Components <- mutate(Tahoe_gage_height_Components,
                      Observed = USGS_gage_processed$gage_height,     
                      Date = USGS_gage_processed$DATE)

# Visualize how the trend maps onto the data
ggplot(Tahoe_gage_height_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab("Gage Height (ft)")

# Visualize how the seasonal cycle maps onto the data
ggplot(Tahoe_gage_height_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab("Gage Height (ft)")
```

#Trend analysis

Monotonic trends are a gradual shift over time that is consistent in direction, for example in response to land use change.
We are going to test monotonic trends. We have no missing gaps in our data. 
**Seasonal Mann-Kendall**: seasonality, non-parametric, no temporal autocorelation, identical distribution. Function: `smk.test` (package: trend)

We are interested in knowing how lake level has changed over time while incorporating the seasonal component. The Seasonal Mann-Kendall assumes no temporal autocorrelation, but we know that daily data is prone to temporal autocorrelation. In this case, we may want to collapse our data down into monthly data so that we can (1) reduce temporal autocorrelation and (2) break down the potential seasonal trend into more interpretable components. 

```{r}
#monthly mean gage height 
Tahoe_gage_height_monthly <- USGS_gage_processed %>%
  mutate(Year = year(DATE),
         Month = month(DATE)) %>%
  group_by(Year, Month) %>%
  summarise(gage_height = mean(gage_height))

Tahoe_gage_height_monthly$Date <- as.Date(paste(Tahoe_gage_height_monthly$Year,
                                                Tahoe_gage_height_monthly$Month,
                                                1, sep = "-"),
                                          format = "%Y-%m-%d")

# Generate time series (smk.test needs ts, not data.frame)
Tahoe_gage_monthly_ts <- ts(Tahoe_gage_height_monthly$gage_height, frequency = 12, 
                            start = c(1957, 10, 1), end = c(2019, 12, 1))

# Run SMK test only on a time series object
Tahoe_gage_height_trend <- smk.test(Tahoe_gage_monthly_ts)

# Inspect results
Tahoe_gage_height_trend
summary(Tahoe_gage_height_trend)
#significant trend over time bc p-value is 2.827e-09 (less than 0.05), z = -5.9414

ggplot(Tahoe_gage_height_monthly, aes(x = Date, y = gage_height)) +
  geom_point() +
  geom_line()

```

If a significant trend was present, we could compute a **Sen's Slope** to quantify that trend (`sens.slope` function in the trend package).

```{r}
#Sen's Slope to quantify trend of lake level over time 


```


